Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-07 article.Rmd
Revised rest of the introduction

But researchers studying historical subjects need to take into account the dynamic relationship between gender and naming practices. If that same researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, Leslie Fay’s gender would get misidentified. And the “Leslie” problem is not just one for historians studying the past. Anyone studying a longer timeframe will also encounter this problem: as of 2012, the average American has a lifespan of nearly seventy-nine years – more than enough time for naming practices to change quite dramatically. (cite World Bank).

Our solution to the “Leslie” problem is to create a library that explicitly takes into account changes in naming practices. The program incorporates this temporal dimension using historical databases to more accurately infer whether “Leslie” was male or female depending on when he or she lived. It not only achieves better accuracy in inferring gender, but it also provides the relative likelihood of a person’s gender for a given range of years: 52% of the babies named “Leslie” born in 1950 were girls and 48% were boys.  Rather than a black box spitting out a stark FEMALE or MALE designation, the program computes a measurement for each gender’s relative “share” of a given name. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of male or female.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is necessary since the large data sets which make this kind of analysis possible were created by state agencies, and for gender (as for race) state agencies have created a limited number of categories which have tremendous definitional power. [CITE e.g. Margot Canady, The Straight State; works on e.g. Virginia’s definition of race. Shane Landrum on birth certificates. James C. Scott on Seeing Like a State or is that obvious to everyone by now?] Nevertheless, this method is suitable for two reasons. First, researchers are likely to use this method in the aggregate rather than for individuals. The assumption that a name implies male or female (and that those are possible categories to be inferred) has a higher validity when applied to large groups of people than it does when applied to individuals. Second, this methods is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying data set so that a technologically competent researcher can interrogate every assumption through the process. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

The remainder of the article is divided into two sections. First, we present the methodology of our approach by comparing it to existing methods and detailing the strengths and weaknesses of the program. Second, we apply the method to a case study in order to demonstrate how humanities scholars might use it by analyzing gender representation within the historical profession.

>>>>

Researchers studying historical subjects, on the other hand, need to take into account changes in naming practices. If that same researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary datasets would misidentify Leslie Fay's gender. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a timeframe longer than a few years, or anyone studying a group whose characteristics do not match the probably unknowable characteristics of the groups that comprise social media datasets will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically.^[CITATION] Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a range of historical dataset suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48% were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of male or female.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large datasets which make this kind of analysis possible for the past were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define.^[CITE e.g. Margot Canady, The Straight State; works on e.g. Virginia’s definition of race. Shane Landrum on birth certificates. James C. Scott on Seeing Like a State or is that obvious to everyone by now?] Nevertheless, this method is suitable for two reasons. First, researchers are likely to use this method in the aggregate rather than for individuals. The assumption that a name implies male or female (and that those are possible categories to be inferred) is more valid when applied to large groups of people than it does when applied to individuals. Second, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption we have made. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use the method. Second, we apply the method to a case study of the historical profession in order to demonstrate the usefulness of the method.
